Ecosystem size and complexity are extrinsic drivers of food chain length in branching networks

### 2022-05-09 rmd/manuscript.Rmd
edit: maintext

Understanding the drivers of food chain length in natural communities has intrigued ecologists since Elton publicized 'food cycles' in the early 20th century. Proposed drivers of food chain length have included productivity, disturbance regime, ecosystem size, and trophic omnivory. However, current theories have largely assumed simple, two-dimensional habitat architectures and may not be adequate to predict food chain length in ecosystems with a complex, branching structure. Here, we develop a spatially explicit theoretical model which provides an integrated framework for understanding variation in food chain length in branching networks. We show independent, positive influences of ecosystem size and complexity on food chain length. However, these effects were contingent upon other proposed drivers, such as the disturbance regime. Our results suggest that ecosystem complexity is an important yet overlooked driver of food chain length in branching ecosystems.

@@

Among the three hypotheses, the ecosystem size hypothesis has been consistently supported in empirical explorations, likely because multiple mechanisms may underlie the positive association [@takimotoEnvironmentalDeterminantsFoodchain2013]. For example, large ecosystems have increased species richness (species-area relationship [@lomolinoEcologyMostGeneral2000]), buffer against disturbances [@saboRoleDischargeVariation2010], and have greater numbers of immigrant sources in the ecosystem with stable metapopulations [@takimotoEnvironmentalDeterminantsFoodchain2013]. Meanwhile, the theoretical basis for the other two hypotheses rests on implicit assumptions of linear food chains, which are often violated in natural systems. As such, the effects of disturbance and resource availability on FCL have been found to be highly variable among ecosystem types and geographic regions [@takimotoEffectsProductivityDisturbance2012a; @diehlIntraguildPreySuffer2001; @takimotoEnvironmentalDeterminantsFoodchain2013; @postLongShortFoodchain2002].

More recently, food web ecologists have begun to recognize that FCL rarely responds to a single driver, but instead is determined by interactions of multiple factors. For example, @wardMechanisticTheoryAquatic2017 showed that FCL was more responsive to increasing productivity in large ecosystems than small ones, whereas FCL was more sensitive to size in low-productivity systems. Such context-dependent responses to extrinsic factors seem to be tightly coupled with the arrangement of trophic links within a food web. For example, omnivory (feeding at multiple trophic levels) is a strong intrinsic driver of FCL [@takimotoEffectsProductivityDisturbance2012a; @postProximateStructuralMechanisms2007; @woottonOmnivoryStabilityFreshwater2017]. When omnivory is absent in three-species communities, the trophic structure assumes a simple linear food chain with a maximum length of three (**Figure 1D**). However, FCL decreases proportionally as the top predator eats lower in the food web as omnivory increases [@wardMechanisticTheoryAquatic2017; @woottonOmnivoryStabilityFreshwater2017]. Theoretical work predicts that food webs are destabilized when omnivory is prevalent, leading to the loss of species and shorter food chains even in productive, large ecosystems. Conversely, weak omnivory stabilizes the food web and increases the persistence of species within a community, particularly at intermediate levels of productivity and high disturbance levels [@woottonOmnivoryStabilityFreshwater2017; @woottonManyWeakInteractions2016]. Due to the context-dependency of FCL to multiple drivers and the potential for interactive effects, a unified framework of multiple food chain drivers is needed to better understand when and where FCL will respond to these extrinsic and intrinsic factors.

@@

We simulated the spatial distribution of disturbance intensity (i.e., proportional mortality in the event of stochastic regional disturbance) as follows. A disturbance value, $m_{h}$, was assigned randomly to upstream terminal patch $h$ as $logit~m_{h} \sim Normal(logit~\mu_m, \sigma_{h}^2)$, where $\mu_m$ is the average of disturbance intensity at headwaters in an ordinary scale. The disturbance values propagate downstream through a spatial autoregressive process as:

@@

where $m_{down}$ and $m_{up}$ are the disturbance values at down- and upstream patches, and the local environmental noise at the downstream patch $\epsilon$ was drawn from $Normal(0, \sigma_{l}^2)$. This autoregressive process resembles a natural disturbance regime, where flood or drought disturbance cascades downstream with some local variation along the channel. At confluence patches, the disturbance value takes a weighted mean based on the relative size of the two contributing branches, $Q$ (the number of upstream contributing patches):

@@

We also mimicked the downstream increase of local carrying capacity [@teruiThreeEcologicalFactors2016; @carraraComplexInteractionDendritic2014; @rahelFishAssemblagesHabitat1991]. Carrying capacity in patch $x$, $K_x$, was assumed to increase as a power function of ecosystem area (i.e., watershed area) [@finlayStreamSizeHuman2011; @koenigEmergentProductivityRegimes2019]:

@@

where $K_{base}$ is the carrying capacity at the upstream terminal patch, $N_{p,up}$ the number of patches upstream, and $z$ is the scaling exponent. We used $K_{base} = 100$ and $z = 0.54$ based on an empirical relationship of gross primary production and watershed area [@finlayStreamSizeHuman2011].

@@

$$
\hat{A}_{P,x}(t) =\left\{\begin{aligned}&s_{0,P} \left[e_{BP} \delta_{x}(t) W_{BP,x}(t) + e_{CP} (1 - \delta_{x}(t)) W_{CP,x}(t) \right] &&\text{if~$W_{B,x}(t) < A_{B,x}(t)$}\\&s_{0,P} \left[e_{BP} \left(\frac{\delta_{x}(t)W_{BP,x}(t)}{W_{B,x}(t)}A_{B,x}\right) + e_{CP} (1 - \delta_{x}(t)) W_{CP,x}(t) \right] &&\text{if~$W_{B,x}(t) > A_{B,x}(t)$}\\\end{aligned}\right.
$$

@@

Before the main simulation, we performed an extensive sensitivity analysis to identify influential parameters (see **Supplementary Text** and **Table S2-S3**). Our sensitivity analysis revealed that ecosystem size and/or complexity effects were sensitive to disturbance, productivity, dispersal, and omnivory. Therefore, we ran the main simulation with `r n_scenario` parameter sets differing in these ecological contexts (two levels of disturbance, two levels of productivity, two levels of dispersal distance, and three levels of omnivory).

We controlled disturbance level with disturbance probability $p_{m}$ (low disturbance, `r min(sim_main_result$p_disturb)`; high disturbance, `r max(sim_main_result$p_disturb)`). We used the identical value of mean disturbance intensity at headwaters ($\mu_{m}=$ `r max(sim_main_result$mean_disturb_source)`), spatial variation in disturbance intensity at headwaters ($\sigma_{h}=$ `r min(sim_main_result$sd_disturb_source)`), and local disturbance variation ($\sigma_{l}=$ `r unique(sim_main_result$sd_disturb_lon)`) regardless of disturbance scenario. Productivity was varied by setting different values of basal species' growth rate ($r_{b}=$ `r unique(sim_main_result$r_b)`). For dispersal, we examined short- and long-distance dispersal scenarios (short-distance dispersal $\theta=$ `r max(sim_main_result$theta)`; long-distance dispersal $\theta=$ `r min(sim_main_result$theta)`), while using constant dispersal probability ($p_{d}=$ `r unique(sim_main_result$p_dispersal)`).

@@

In ecosystems with a low disturbance regime, omnivory largely determined FCL (colored lines in **Figure 2**), with no-omnivory food chains ($P$ only consumes $C$) having the longest FCL, and strong omnivory ($P$ consumes very little $C$) having the shortest FCL. The primary mechanism for the decreased FCL was different in low and high productivity scenarios. With low productivity, strong omnivory caused shorter FCL because it intensifies competition between consumer and predator species. We can see this effect in the decreased occupancies of consumer and predator species in a low productivity scenario (most left column in **Figure S1-S4**). However, when productivity was high, the occupancies of both species were \~1.0, indicating that the structural constraint of strong omnivory made FCL shorter (second left column in **Figure S1-S4**). Productivity played a secondary role in increasing FCL, particularly when omnivory was strong (compare left and right panels in **Figure 2**). Dispersal had no apparent effects on FCL (compare top and bottom panels in **Figure 2**). These trends were consistent across dispersal scenarios (**Figure 2**).

Although the effects of ecosystem properties on FCL were minimal, there were noticeable effects on the occupancies of constituent species in some scenarios. For example, the occupancies of $C$ and $P$ increased with ecosystem size when omnivory was strong (see the most left column in **Figures S1 and S2**). However, these changes did not translate into FCL because $C$ and $P$ occupy a similar trophic position (**Figure 1D**). Meanwhile, the occupancy of $P$ decreased with ecosystem complexity when omnivory was strong, but FCL remained almost the same as the increase of $C$ could compensate for the loss of $P$ (**Figure S3 and S4**).

@@

On average, increased disturbance frequency decreased FCL, and omnivory had similar effects on FCL (**Figures 3**). More importantly, a high disturbance regime produced different responses of FCL to ecosystem properties. FCL increased across gradients of ecosystem size and complexity under all productivity, omnivory, and dispersal scenarios (**Figure 3**). The effect of ecosystem complexity was generally clearer than that of ecosystem size. It is important to note that the positive relationships emerged in a high disturbance regime because smaller and simpler ecosystems were disproportionately vulnerable to increased disturbance frequency (compare far left in **Figures 2 and 3**).

@@

Different processes drove the positive effects of ecosystem size and complexity on FCL when strong omnivory exists in low-productivity systems. In this scenario, the occupancies of $B$ and $C$ decreased at larger ecosystem sizes, while $P$ increased its occupancy (third column in **Figure S1 and S2**). Therefore, $P$ compensated for the loss of $B$ and $C$ to maintain the positive relationship between FCL and ecosystem size. In contrast, the increased occupancy of $C$ maintained the positive relationship between FCL and ecosystem complexity by offsetting the loss of $P$.

@@

The ecosystem size hypothesis of FCL has received support from both theoretical and empirical research [@vanderzandenPatternsFoodChain1999; @postEcosystemSizeDetermines2000; @takimotoEffectsProductivityDisturbance2012a], although empirical work so far has rarely elaborated on the underlying drivers [but see @saboRoleDischargeVariation2010; @wardMechanisticTheoryAquatic2017]. In contrast, ecosystem complexity has not been incorporated in theoretical explorations of FCL to date. Here, we show that FCL responds positively to increasing ecosystem size and complexity. This supports our hypothesis that ecosystem complexity may represent an overlooked yet important extrinsic control of FCL. However, the relationship between FCL and ecosystem complexity varied based on additional extrinsic and intrinsic factors, supporting the emerging paradigm that there is no single driver of FCL in ecosystems [@mchughDualInfluencesEcosystem2010; @wardMechanisticTheoryAquatic2017]. Although we focus on river networks here, we expect our results to be transferable to other branching structures, such as mountain ranges and cave systems.

While increased disturbance frequency consistently reduced FCL as predicted by the dynamical stability hypothesis [@pimmNumberTrophicLevels1977], more important findings in this study were evident from their synergistic effects on ecosystem structure. Our low disturbance regime scenario can be viewed as habitat networks having nearly homogeneous environments, in which long food chains were generally supported independently of ecosystem size and complexity. Consequently, the food web structure largely controlled FCL. However, under high disturbance regimes, only large and complex networks could provide diverse and heterogeneous habitats with potential refugia, allowing species to thrive and maintain long food chains. While empirical studies rarely looked at the influence of ecosystem structure on FCL, several studies provide some support for this theoretical prediction. For example, @sullivanEcosystemStructureEmerges2015 reported FCL is controlled primarily by the number of tributary confluences in Idaho streams. Similarly, it has been shown that the merging of major tributaries lengthened FCL in the Colorado River, especially during a post-monsoon turbid water phase [@saboPulsedFlowsTributary2018]. More empirical research is needed to confirm our theoretical predictions.

@@

Our findings add an important distinction in the discussion of the suitability of assigning protection or conservation status to regions. Current conservation schemes generally emphasize size, aiming to protect the largest areas possible. While this is certainly a useful attribute generally, designating networks with a high degree of variability, or many small patches, can be more beneficial than a large, homogenous area [@fahrigWhySeveralSmall2020]. Accounting for ecosystem complexity is particularly suited to ecosystem architectures in which branches have variable environmental conditions [@koningNetworkGrassrootsReserves2020; @teruiEmergentDualScaling2021]. Conservation and management activities should focus on activities that increase the habitat heterogeneity and connectivity of suitable habitat patches to increase resilience to predicted increases in disturbance frequency and magnitude.

@@

JP developed the theoretical model of trophic dynamics, ran the simulations, and made the figures with input from AT. JP wrote the first draft, while AT and JF contributed significantly to the final version.

>>>>

Understanding the drivers of food chain length in natural communities has intrigued ecologists since Elton publicized 'food cycles' in the early 20th century. Proposed drivers of food chain length have included productivity, disturbance regime, ecosystem size, and trophic omnivory. However, current theories have largely assumed simple, two-dimensional habitat architectures and may not be adequate to predict food chain length in ecosystems with a complex, branching structure. Here, we develop a spatially explicit theoretical model which provides an integrated framework for understanding variation in food chain length in branching networks. We show independent, positive influences of ecosystem size and complexity on food chain length. However, these effects were contingent upon other drivers. In particular, the effects of ecosystem size and complexity appeared more clearly in high disturbance regimes. Our results suggest that ecosystem complexity is an important yet overlooked driver of food chain length in branching ecosystems.

@@

Among the three hypotheses, the ecosystem size hypothesis has been consistently supported in empirical explorations, likely because it encompasses multiple mechanisms [@takimotoEnvironmentalDeterminantsFoodchain2013]. For example, large ecosystems have increased species richness (species-area relationship [@lomolinoEcologyMostGeneral2000]), are buffered against disturbances [@saboRoleDischargeVariation2010], and have greater numbers of immigrant sources in the ecosystem with stable metapopulations [@takimotoEnvironmentalDeterminantsFoodchain2013]. Meanwhile, the theoretical basis for the other two hypotheses rests on implicit assumptions of linear food chains, which are often violated in natural systems. As such, the effects of disturbance and resource availability on FCL have been found to be highly variable among ecosystem types and geographic regions [@takimotoEffectsProductivityDisturbance2012a; @diehlIntraguildPreySuffer2001; @takimotoEnvironmentalDeterminantsFoodchain2013; @postLongShortFoodchain2002].

More recently, food web ecologists have begun to recognize that FCL rarely responds to a single driver, but instead is determined by interactions of multiple factors. For example, @wardMechanisticTheoryAquatic2017 showed that FCL was more responsive to increasing productivity in large ecosystems than small ones, whereas FCL was more sensitive to size in low-productivity systems. Such context-dependent responses to extrinsic factors seem to be tightly coupled with the arrangement of trophic links within a food web. For example, omnivory (feeding at multiple trophic levels) is a strong intrinsic driver of FCL [@takimotoEffectsProductivityDisturbance2012a; @postProximateStructuralMechanisms2007; @woottonOmnivoryStabilityFreshwater2017]. When omnivory is absent in three-species communities, the trophic structure assumes a simple linear food chain with a maximum length of three (**Figure 1D**). However, FCL decreases proportionally with increasing omnivory as the top predator eats lower in the food web [@wardMechanisticTheoryAquatic2017; @woottonOmnivoryStabilityFreshwater2017]. Theoretical work predicts that food webs are destabilized when omnivory is too strong, leading to the loss of species and shorter food chains [@woottonOmnivoryStabilityFreshwater2017]. Conversely, weak omnivory stabilizes the food web and increases the persistence of species within a community, particularly at intermediate levels of productivity and high disturbance levels [@woottonOmnivoryStabilityFreshwater2017; @woottonManyWeakInteractions2016]. Due to the context-dependency of FCL to multiple drivers and the potential for interactive effects, a unified framework of multiple FCL drivers is needed to better understand when and where FCL will respond to these extrinsic and intrinsic factors.

@@

In our simulation, disturbance occurs simultaneously with probability $p_m$ but habitat patches vary in their vulnerability. We simulated the spatial distribution of disturbance intensity (i.e., proportional mortality in the event of stochastic regional disturbance) as follows. A disturbance value, $m_{h}$, was assigned randomly to upstream terminal patch $h$ as $logit~m_{h} \sim Normal(logit~\mu_m, \sigma_{h}^2)$, where $\mu_m$ is the average of disturbance intensity at headwaters in an ordinary scale. The disturbance values propagate downstream through a spatial autoregressive process as:

@@

where $m_{down}$ and $m_{up}$ are the disturbance values at down- and upstream patches, and the local environmental noise at the downstream patch $\epsilon$ was drawn from $Normal(0, \sigma_{l}^2)$. This autoregressive process resembles a natural disturbance regime, where flood or drought disturbance cascades downstream with some local variation along the channel [@nakamuraDisturbanceRegimesStream2000]. At confluence patches, the disturbance value takes a weighted mean based on the relative size of the two contributing branches, $Q$ (the number of upstream contributing patches):

@@

We also mimicked the downstream increase of local carrying capacity reflecting downstream increases in algal resource availability [@teruiThreeEcologicalFactors2016; @carraraComplexInteractionDendritic2014; @rahelFishAssemblagesHabitat1991]. Carrying capacity in patch $x$, $K_x$, was assumed to increase as a power function of ecosystem area (i.e., watershed area) [@finlayStreamSizeHuman2011; @koenigEmergentProductivityRegimes2019]:

@@

where $K_{base}$ is the carrying capacity at the upstream terminal patch, $N_{p,up}$ the number of patches upstream, and $z$ the scaling exponent. We used $K_{base} = 100$ and $z = 0.54$ based on an empirical relationship of gross primary production and watershed area [@finlayStreamSizeHuman2011].

@@

$$
\hat{A}_{P,x}(t) =\left\{\begin{aligned}&s_{0,P} \left[e_{BP} \delta_{x}(t) W_{BP,x}(t) + e_{CP} (1 - \delta_{x}(t)) W_{CP,x}(t) \right] &&\text{if~$W_{B,x}(t) < A_{B,x}(t)$}\\&s_{0,P} \left[e_{BP} \delta_{x}(t)\left(\frac{W_{BP,x}(t)}{W_{B,x}(t)}A_{B,x}\right) + e_{CP} (1 - \delta_{x}(t)) W_{CP,x}(t) \right] &&\text{if~$W_{B,x}(t) > A_{B,x}(t)$}\\\end{aligned}\right.
$$

@@

Before the main simulation, we performed an extensive sensitivity analysis to identify influential parameters (see **Supplementary Text** and **Table S1-S3**). Our sensitivity analysis revealed that ecosystem size and/or complexity effects were sensitive to disturbance, productivity, dispersal, and omnivory. Therefore, we ran the main simulation with `r n_scenario` parameter sets differing in these ecological contexts (two levels of disturbance, two levels of productivity, two levels of dispersal distance, and three levels of omnivory) and parameter values were summarized in **Table S1**.

We controlled disturbance level with disturbance probability $p_{m}$ (low disturbance, `r min(sim_main_result$p_disturb)`; high disturbance, `r max(sim_main_result$p_disturb)`). We used the identical value of mean disturbance intensity at headwaters ($\mu_{m}=$ `r max(sim_main_result$mean_disturb_source)`), spatial variation in disturbance intensity at headwaters ($\sigma_{h}=$ `r max(sim_main_result$sd_disturb_source)`), and local disturbance variation ($\sigma_{l}=$ `r min(sim_main_result$sd_disturb_lon)`) regardless of disturbance scenario. This setup ($\sigma_h > \sigma_l$) produces high spatial variation among tributaries with strong within-tributary downstream correlation, a typical pattern in real river networks [@nakamuraDisturbanceRegimesStream2000; @bendaNetworkDynamicsHypothesis2004]. Productivity was varied by setting different values of basal species' growth rate ($r_{b}=$ `r unique(sim_main_result$r_b)`). For dispersal, we examined short- and long-distance dispersal scenarios (short-distance dispersal $\theta=$ `r max(sim_main_result$theta)`; long-distance dispersal $\theta=$ `r min(sim_main_result$theta)`) with constant dispersal probability ($p_{d}=$ `r unique(sim_main_result$p_dispersal)`).

@@

In ecosystems with a low disturbance regime, omnivory largely determined FCL (colored lines in **Figure 2**), with no-omnivory food chains ($P$ only consumes $C$) having the longest FCL, and strong omnivory ($P$ consumes very little $C$) having the shortest FCL. The primary mechanism for the decreased FCL was different in low and high productivity scenarios. With low productivity, strong omnivory caused shorter FCL because it intensified competition between consumer and predator species. We can see this effect in the decreased occupancies of consumer and predator species in a low productivity scenario (most left column in **Figure S1-S4**). However, when productivity was high, the occupancies of both species were \~1.0, indicating that the structural constraint of strong omnivory made FCL shorter (second column from left in **Figure S1-S4**). Productivity played a secondary role in increasing FCL, particularly when omnivory was strong (compare left and right panels in **Figure 2A and 2B**). Dispersal had no apparent effects on FCL (compare top and bottom panels in **Figure 2A and 2B**).

Although the effects of ecosystem properties on FCL were minimal, there were noticeable effects on the occupancies of constituent species in some scenarios. For example, the occupancies of $C$ and $P$ increased with ecosystem size when omnivory was strong (see bottom two rows in left most column in **Figures S1 and S2**). However, these changes did not translate into FCL because $C$ and $P$ occupy a similar trophic position (**Figure 1D**). Meanwhile, the occupancy of $P$ decreased with ecosystem complexity when omnivory was strong, but FCL remained almost the same as the increase of $C$ could compensate for the loss of $P$ (**Figure S3 and S4**).

@@

On average, increased disturbance frequency decreased FCL, and omnivory had similar effects on FCL (**Figure 3**). More importantly, a high disturbance regime produced different responses of FCL to ecosystem properties. FCL increased across gradients of ecosystem size and complexity under all productivity, omnivory, and dispersal scenarios (**Figure 3**). The effect of ecosystem complexity was generally clearer than that of ecosystem size. It is important to note that the positive relationships emerged in a high disturbance regime because smaller and simpler ecosystems were disproportionately vulnerable to increased disturbance frequency (compare far left along the x-axis in each panel of **Figures 2 and 3**).

@@

Different processes drove the positive effects of ecosystem size and complexity on FCL when strong omnivory exists in low-productivity systems. In this scenario, the occupancies of $B$ and $C$ decreased at larger ecosystem sizes, while $P$ increased its occupancy (third column from left in **Figure S1 and S2**). Therefore, $P$ compensated for the loss of $B$ and $C$ to maintain the positive relationship between FCL and ecosystem size. In contrast, the increased occupancy of $C$ maintained the positive relationship between FCL and ecosystem complexity by offsetting the loss of $P$.

@@

The ecosystem size hypothesis of FCL has received support from both theoretical and empirical research [@vanderzandenPatternsFoodChain1999; @postEcosystemSizeDetermines2000; @takimotoEffectsProductivityDisturbance2012a], although empirical work so far has rarely elaborated on the underlying drivers [but see @saboRoleDischargeVariation2010; @wardMechanisticTheoryAquatic2017]. In contrast, ecosystem complexity has not been incorporated in theoretical explorations of FCL to date. Here, we show that FCL responds positively to increasing ecosystem size and complexity, here assessed as the number of patches and branching probability in river networks. This supports our hypothesis that ecosystem complexity may represent an overlooked yet important extrinsic control of FCL. However, the relationship between FCL and ecosystem complexity varied with other extrinsic and intrinsic factors, supporting the emerging paradigm that FCL is regulated by multiple factors [@mchughDualInfluencesEcosystem2010; @wardMechanisticTheoryAquatic2017]. Although we focus on river networks here, we expect our results to be transferable to other systems with branching structures, such as mountain ranges and cave systems. Our findings should be relevant to non-branching systems, where different metrics of ecosystem complexity would be necessary.

While increased disturbance frequency consistently reduced FCL as predicted by the dynamical stability hypothesis [@pimmNumberTrophicLevels1977], more important findings in this study were evident from their synergistic effects with ecosystem structure. Our low disturbance regime scenario can be viewed as habitat networks having nearly homogeneous environments, in which long food chains were generally supported independently of ecosystem size and complexity. Consequently, the food web structure largely controlled FCL. However, under high disturbance regimes, only large and complex networks could provide diverse and heterogeneous habitats with potential refugia, allowing species to thrive and maintain long food chains. While empirical studies rarely looked at the influence of ecosystem structure on FCL, several studies provide some support for this theoretical prediction. For example, @sullivanEcosystemStructureEmerges2015 reported FCL is controlled primarily by the number of tributary confluences in Idaho streams. Similarly, it has been shown that the merging of major tributaries lengthened FCL in the Colorado River, especially during a post-monsoon turbid water phase [@saboPulsedFlowsTributary2018]. More empirical research is needed to confirm our theoretical predictions.

@@

Our findings add an important distinction to work to assess and prioritize protection or conservation status within human impacted regions. Classical conservation schemes emphasize ecosystem size, aiming to protect the largest areas possible [@diamondIslandDilemmaLessons1975]. While this is certainly a useful attribute generally, designating networks with a high degree of variability, or many small patches, can be more beneficial than a large, homogeneous area [@fahrigWhySeveralSmall2020]. Accounting for ecosystem complexity is particularly suited to ecosystem architectures in which branches have variable environmental conditions [@koningNetworkGrassrootsReserves2020; @teruiEmergentDualScaling2021]. As suggested by modern conservation initiatives, conservation and management activities should focus on activities that increase the habitat heterogeneity and connectivity of suitable habitat patches to increase resilience to predicted increases in disturbance frequency and magnitude [@schindlerPopulationDiversityPortfolio2010; @schindlerPortfolioConceptEcology2015a; @andersonPortfolioConservationMetapopulations2015].

@@

JP developed the theoretical model of trophic dynamics with input from AT. JP and AT ran the simulations, and made the figures. JP wrote the first draft, while AT and JF contributed significantly to the final version.
